---
title: "Adding first or last events"
output:
  html_document:
    pandoc_args: [
      "--number-offset=1,0"
      ]
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{event}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Introduction

The event functions identify which event occurs first or last within a time
window. They can search either a cohort table or a concept set:

| Event source | Return the date | Return days relative to the index date |
|:--|:--|:--|
| Cohort table | `addCohortEventDate()` | `addCohortEventDays()` |
| Concept set | `addConceptEventDate()` | `addConceptEventDays()` |

Each function adds two columns for every window:

- an `event` column containing the name of the selected event; and
- a `date` or `days` column containing when that event occurred.

The cohort tables searched by the event functions below are created with `CohortConstructor`. PatientProfiles then searches these valid cohort tables but does not create or modify cohort definitions; those tasks can be handled by `CohortConstructor` or other dedicated OMOP cohort-generation packages.

Unlike the intersection functions, the event functions also return the
applicable boundary when no target event occurs in the requested window. This
makes it possible to distinguish an observed event, an explicit censoring
boundary, and the end of observation.

We will use the *GiBleed* dataset to illustrate this functionality:

```{r}
library(PatientProfiles)
library(dplyr)
library(omock)
library(CohortConstructor)
library(CodelistGenerator)

cdm <- mockCdmFromDataset(datasetName = "GiBleed", source = "duckdb")
```

Let's create some simple cohorts:

```{r}
codelist <- getDrugIngredientCodes(
  cdm = cdm,
  name = "acetaminophen",
  nameStyle = "{concept_name}"
)
cdm$my_cohort <- conceptCohort(
  cdm = cdm,
  conceptSet = codelist,
  name = "my_cohort"
) |>
  requireIsFirstEntry()
codelist <- list(osteoarthritis = 80180L, diverticular_disease = 4266809L)
cdm$conditions <- conceptCohort(
  cdm = cdm,
  conceptSet = codelist,
  name = "conditions"
)
```

## Finding a cohort event

`addCohortEventDays()` searches the cohorts in `targetCohortTable`. In this
example we find the first event on or after each record's cohort start date and
up to 365 days later.

```{r}
cohort_event_days <- cdm$my_cohort |>
  addCohortEventDays(
    targetCohortTable = "conditions",
    indexDate = "cohort_start_date",
    order = "first",
    window = list(next_year = c(0, 365))
  )

cohort_event_days |>
  glimpse()
```

The `days_next_year` value is relative to `cohort_start_date`: positive values
are after the index date, zero is the index date, and negative values are before
it.

We can obtain a quick summary of the result:

```{r}
cohort_event_days |>
  group_by(event_next_year) |>
  summarise(
    n = n(),
    median_days = median(days_next_year)
  ) |>
  collect()
```

- **censor** means that no event was found before the window or censoring
  boundary;
- **end_of_observation** means that no event was found before the relevant
  observation period boundary; and
- any other value is the name of the event that was found.

To find the nearest previous event, use a window ending on the index date and
`order = "last"`. Here, `"last"` selects the latest event in the window. Because
the complete window is in the past, this is the event nearest the index date.

```{r}
cohort_event_date <- cdm$my_cohort |>
  addCohortEventDate(
    targetCohortTable = "conditions",
    indexDate = "cohort_start_date",
    order = "last",
    window = list(previous_year = c(-365, 0))
  )

cohort_event_date |>
  glimpse()
```

Use `targetCohortId` to restrict the search to particular cohorts.
`targetDate` determines which date in the target cohort table represents the
event and defaults to `cohort_start_date`.

## Finding a concept event

The concept event functions search records associated with one or more concept
sets across the OMOP clinical tables. The names of the concept sets become the
event names.

```{r}
concept_event_days <- cdm$my_cohort |>
  addConceptEventDays(
    conceptSet = codelist,
    order = "first",
    window = list(next_event = c(0, Inf))
  )

concept_event_days |>
  glimpse()
```

For concept events, `targetDate` can be either `event_start_date`, the default,
or `event_end_date`. The concept set is validated against the vocabulary in the
CDM before the clinical records are searched.

## First, last, and censoring

The `window` defines the period to search, while `order` determines which event
within that period is returned:

- `order = "first"` selects the earliest event in the window;
- `order = "last"` selects the latest event in the window;
- a window such as `c(-365, 0)` searches before `indexDate`, while
  `c(0, 365)` searches after it; and
- both window endpoints are included.

The search is always restricted to the observation period containing the index
date. For a forward search, a column supplied through `censorDate` can shorten
the available follow-up. If no event occurs before the applicable boundary, the
event column contains:

- `"end_of_observation"` when the observation period boundary is reached; or
- `"censor"` when `censorDate` or the requested window boundary is reached.

The accompanying date or days value is the applicable boundary.

For a forward search, the boundary is the earliest of the upper window
boundary, `censorDate`, and the observation period end. For a backward search,
the boundary is the latest of the lower window boundary and the observation
period start.

An event on the boundary is returned as the event itself. It is not combined
with `"censor"` or `"end_of_observation"`. If the index date is outside
observation, both added values are missing.

## Events occurring on the same date

Several cohorts or concept sets can occur on the selected date. The
`multipleEvents` argument controls how these ties are represented:

- `NULL`, the default, returns the first matching event in the original order;
- `TRUE` returns all matching event names in alphabetical order, separated by
  `"; "`; and
- a character vector defines a priority order and returns the first matching
  name. Event names omitted from the vector follow in alphabetical order.

Use `NULL` to follow the original order of the concept set:

```{r}
cdm$my_cohort |>
  addConceptEventDays(
    conceptSet = codelist,
    order = "first",
    window = list("next" = c(0, Inf)),
    multipleEvents = NULL
  ) |>
  group_by(event_next) |>
  summarise(n = n()) |>
  collect()
```

Use `TRUE` to combine all events that occur on the selected date:

```{r}
cdm$my_cohort |>
  addConceptEventDays(
    conceptSet = codelist,
    order = "first",
    window = list("next" = c(0, Inf)),
    multipleEvents = TRUE
  ) |>
  group_by(event_next) |>
  summarise(n = n()) |>
  collect()
```

Alternatively, provide an explicit priority order. Here,
`osteoarthritis` is selected before `diverticular_disease` when they occur on
the same date:

```{r}
cdm$my_cohort |>
  addConceptEventDays(
    conceptSet = codelist,
    order = "first",
    window = list("next" = c(0, Inf)),
    multipleEvents = c("osteoarthritis", "diverticular_disease")
  ) |>
  group_by(event_next) |>
  summarise(n = n()) |>
  collect()
```

In this example, 390 cohort records have both *osteoarthritis* and
*diverticular disease* on the selected date. The three calls above show how
`multipleEvents` can resolve these ties using the original order, a combined
label, or an explicit priority.

## Multiple windows and column names

More than one window can be supplied in a named list. The default
`nameStyle = "{value}_{window_name}"` uses `{value}` for `event`, `date`, or
`days`, and `{window_name}` for the name of the window.

```{r}
events_in_two_directions <- cdm$my_cohort |>
  addCohortEventDays(
    targetCohortTable = "conditions",
    order = "first",
    window = list(
      before = c(-365, 0),
      after = c(0, 365)
    )
  )

events_in_two_directions |>
  glimpse()
```

The same value of `order` applies to every supplied window. If different
windows require different values of `order`, use separate calls.

`nameStyle` must contain `{value}` so that the event and its date or days have
different names. It must also produce a unique name for every output column.
